---
title: "Isolating data with dplyr"
subtitle: "also: pipes! %>%"
author: "Byron C. Jaeger"
date: "Last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/intro_2R_hex.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: inverse, center, middle

```{r, include = FALSE}

library(magrittr)
library(dplyr)
library(readr)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 5.5,
  dpi = 72 * 5
)

```

# `select`, `filter`, `arrange`

---

## Data:

```{r, echo = FALSE}

nhanes <- read_rds('data/NHANES_analysis_pre_exclusions.rds') %>% 
  select(exam, age, sex, bp_sys_mmhg, n_msr_sbp, bp_meds) %>% 
  tidyr::drop_na() %>% 
  group_by(exam) %>% 
  slice(1:2) %>% 
  ungroup()

nhanes

```
---
class: center, middle

# but first...

---

# Program a story

Here is the legendary tale of Little Bunny Foo Foo.

> Little bunny Foo Foo <br/>
> Went hopping through the forest <br/>
> Scooping up the field mice <br/>
> And bopping them on the head 

--

How would we go about programming this in R?

```{r, eval = FALSE}

foo_foo <- little_bunny()

# now what?

```

---

# New objects

We could save each intermediate step as a new object:

```{r, eval = FALSE}


foo_foo <- little_bunny() # Little bunny Foo Foo

foo_foo_1 <- hop(foo_foo, through = forest) # Went hopping through the forest

foo_foo_2 <- scoop(foo_foo_1, up = field_mice) # Scooping up the field mice

foo_foo_3 <- bop(foo_foo_2, on = head) # And bopping them on the head

```

Problems: 

- The code is cluttered with unimportant names

- You have to carefully increment the suffix on each line.

---

# Overwrite the original

We could overwrite the original object:

```{r, eval = FALSE}

foo_foo <- little_bunny() # Little bunny Foo Foo

foo_foo <- hop(foo_foo, through = forest) # Went hopping through the forest

foo_foo <- scoop(foo_foo, up = field_mice) # Scooping up the field mice

foo_foo <- bop(foo_foo, on = head) # And bopping them on the head


```

less typing (and less likely to make a mistake), but

- Debugging is painful: if you make a mistake you’ll need to re-run the complete pipeline from the beginning.

- The repetition of the object being transformed (we’ve written foo_foo seven times!) obscures what’s changing on each line.

---

# Function composition

Abandon assignment and just string the function calls together:

```{r, eval = FALSE}

bop(
  scoop(
    hop(
      foo_foo, # Little bunny Foo Foo
      through = forest # Went hopping through the forest
    ),
    up = field_mice # Scooping up the field mice
  ), 
  on = head # And bopping them on the head
)
```

Nobody wants to read this. Let's move on.

---

# Use the pipe!

Last, we could use the pipe:

```{r, eval = FALSE}

foo_foo %>%                  # Little bunny Foo Foo
  hop(through = forest) %>%  # Went hopping through the forest
  scoop(up = field_mice) %>% # Scooping up the field mice
  bop(on = head)             # And bopping them on the head

```

--

Pros

- Focusses on verbs, not nouns.

- You can read this series of function compositions like it’s a set of imperative actions: Foo Foo hops, then scoops, then bops.

---

# Use the pipe!

Last, we could use the pipe:

```{r, eval = FALSE}

foo_foo %>%                  # Little bunny Foo Foo
  hop(through = forest) %>%  # Went hopping through the forest
  scoop(up = field_mice) %>% # Scooping up the field mice
  bop(on = head)             # And bopping them on the head

```

Cons

- If you’ve never seen `%>%` before, you’ll have no idea what this code does. 

Fortunately, most people pick up the idea very quickly, so when you share your code with others who aren’t familiar with the pipe, you can easily teach them.

---

# What's a pipe?

Put simply, a pipe takes the object on the left hand side and makes it the first argument for the function on the right hand side. 

```{r}

x <- 1:10

# take the mean of x
mean(x)

# take x, and then find it's mean
x %>% mean()

```


---

# What's a pipe?

More broadly, these two code snippets equivalent.

```{r, eval = FALSE}

# use the function
cool_function(main_arg = x, other_arg = y, last_arg = z)

# start with x and then use the function
x %>% 
  cool_function(other_arg = y, last_arg = z)

```

---

# What's a pipe?

More broadly, these two code snippets equivalent. 

Also, one snippet is much more readable.

```{r, eval = FALSE}

# use the function
another_cool_one(
  cool_function(main_arg = x, other_arg = y, last_arg = z),
  another_arg = w
)


# start with x and then use the function
x %>% 
  cool_function(other_arg = y, last_arg = z) %>% 
  another_cool_one(another_arg = w)

# stylistic tip: end lines at the %>% 

```

---

# When _not_ to pipe

1. Your pipes are longer than (say) five steps. In that case, create intermediate objects with meaningful names.

    + make debugging easier, because you can more easily check the intermediate results 
    + makes it easier to understand your code, because the variable names can help communicate intent.

--

2. You have multiple inputs or outputs. If there isn’t one primary object being transformed, but two or more objects being combined together, don’t use the pipe.

--

3. You are starting to think about a directed graph with a complex dependency structure. Pipes are fundamentally linear and expressing complex relationships with them will typically yield confusing code.

---
class: inverse, center, middle

# Examples

---
class: center, middle

How many males were taking medications to lower blood pressure?

---

```{r}

males <- filter(nhanes, sex == 'Male') #<<
males_bp_meds <- filter(males, bp_meds == 'Yes')
answer <- nrow(males_bp_meds)

```

```{r, echo = FALSE}
males
```

---

```{r}

males <- filter(nhanes, sex == 'Male') 
males_bp_meds <- filter(males, bp_meds == 'Yes') #<<
answer <- nrow(males_bp_meds)

```

```{r, echo = FALSE}
males_bp_meds
```

---

```{r}

males <- filter(nhanes, sex == 'Male') 
males_bp_meds <- filter(males, bp_meds == 'Yes') 
answer <- nrow(males_bp_meds) #<<

```

```{r, echo = FALSE}
answer
```

---

# Pipe solution:

```{r}

answer <- nhanes %>% 
  filter(sex == 'Male', bp_meds == 'Yes') %>% 
  nrow()

answer

```

To make `answer`, I started with `nhanes`, THEN

- I filtered it to contain only males on BP meds, THEN

- I counted how many rows were left over.

---
class: center, middle

What was the highest systolic blood pressure for males? Females?

---

```{r, eval = FALSE}

nhanes %>% #<<
  filter(sex == 'Male') %>% 
  arrange(desc(bp_sys_mmhg)) %>% 
  select(bp_sys_mmhg) %>% 
  slice(1) %>% 
  as.numeric()

```

```{r, echo = FALSE}
nhanes
```

---

```{r, eval = FALSE}

nhanes %>% 
  filter(sex == 'Male') %>% #<<
  arrange(desc(bp_sys_mmhg)) %>% 
  select(bp_sys_mmhg) %>% 
  slice(1) %>% 
  as.numeric()

```

```{r, echo = FALSE}
nhanes %>%
  filter(sex == 'Male')
```

---

```{r, eval = FALSE}

nhanes %>% 
  filter(sex == 'Male') %>% 
  arrange(desc(bp_sys_mmhg)) %>% #<<
  select(bp_sys_mmhg) %>% 
  slice(1) %>% 
  as.numeric()

```

```{r, echo = FALSE}
nhanes %>%
  filter(sex == 'Male') %>% 
  arrange(desc(bp_sys_mmhg)) 
```

---

```{r, eval = FALSE}

nhanes %>% 
  filter(sex == 'Male') %>% 
  arrange(desc(bp_sys_mmhg)) %>% 
  select(bp_sys_mmhg) %>% #<<
  slice(1) %>% 
  as.numeric()

```

```{r, echo = FALSE}
nhanes %>%
  filter(sex == 'Male') %>% 
  arrange(desc(bp_sys_mmhg)) %>% 
  select(bp_sys_mmhg)
```

---

```{r, eval = FALSE}

nhanes %>% 
  filter(sex == 'Male') %>% 
  arrange(desc(bp_sys_mmhg)) %>% 
  select(bp_sys_mmhg) %>% 
  slice(1) %>% #<<
  as.numeric()

```

```{r, echo = FALSE}
nhanes %>%
  filter(sex == 'Male') %>% 
  arrange(desc(bp_sys_mmhg)) %>% 
  select(bp_sys_mmhg) %>% 
  slice(1)
```

---

```{r, eval = FALSE}

nhanes %>%
  filter(sex == 'Male') %>% 
  arrange(desc(bp_sys_mmhg)) %>% 
  select(bp_sys_mmhg) %>% 
  slice(1) %>% 
  as.numeric() #<<

```

```{r, echo = FALSE}
nhanes %>%
  filter(sex == 'Male') %>% 
  arrange(desc(bp_sys_mmhg)) %>% 
  select(bp_sys_mmhg) %>% 
  slice(1) %>% 
  as.numeric()
```

---

# Your turn

- Open up the [Isolate data](https://rstudio.cloud/spaces/15174/project/1096224) project in Rstudio.cloud 

- Start reading the exercises.Rmd file

- Discuss problems among teams!


